---
title: "Numbers in engineering format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Numbers in engineering format}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(
  echo = TRUE, # varies from one Rmd to another
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center",
  fig.path = "../man/figures/engr-format-"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
options(tibble.print_min = 8L, tibble.print_max = 8L)
```

## Introduction

Packages used in this vignette. 

```{r message = FALSE}
library(tidyverse)
library(knitr)
library(docxtools)
```

This vignette demonstrates the use of two functions from the *docxtools* package: 

- `format_engr()` for formatting numbers in engineering notation  
- `align_pander()` for aligning table columns using a simple pander table style 

The primary goal of `format_engr()` is to present numeric variables in a data frame in *engineering format*, that is, scientific notation with exponents that are multiples of 3. Compare: 

|syntax         |expression           |
|---------------|---------------------|
|computer       |$1.011E+5$           | 
|mathematical   |$1.011\times10^{5}$  |
|engineering    |$101.1\times10^{3}$  |


## format_engr()

This example uses a small data set, `density`, included with docxtools, with temperature in K, pressure in Pa, the gas constant in J kg^-1^K^-1^, and density in kg m^-3^.   

```{r}
density
```

Four of the variables are numeric. The `date` variable is of type "double" but class "Date", so it is not reformatted. 

```{r}
map_chr(density, class)
map_chr(density, typeof)
```

Usage is `format_engr(x, sigdig = NULL, ambig_0_adj = FALSE)`. The function returns a data frame with all numeric values reformatted as character strings in engineering format with math delimiters `$...$`.  

```{r}
density_engr <- format_engr(density)
density_engr
```

The formerly numeric variables are now characters.  Non-numeric variables are returned unaltered.

```{r}
map_chr(density_engr, class)
```

The math formatting is applied when the data frame is printed in the output document.  For example, we can use `knitr::kable()` to print the formatted data.  

```{r}
kable(density_engr)
```

The function is compatible with the pipe operator. 

```{r}
density_engr <- density %>%
  format_engr()
kable(density_engr)
```

Comments: 

- as illustrated by the temperature column, scientific notation is omitted when the exponent is 0, 1, or 2 
- trailing zeros after a decimal point are significant 

## Significant digits

`format_engr()` has three arguments: 

- `x`, a data frame with at least one numerical variable. 
- `sigdig`, an optional vector of significant digits. Default is 4. 
- `ambig_0_adj`, an optional logical to adjust the notation in the event of ambiguous trailing zeros. Default is FALSE. 

The `sigdig` argument can be a single value, applied to all numeric columns. 

```{r}
density_engr <- format_engr(density, sigdig = 3)
kable(density_engr)
```

Alternatively, significant digits can be assigned to every numeric column. A zero returns the variable in its original form.   

```{r}
density_engr <- format_engr(density, sigdig = c(5, 4, 0, 5))
kable(density_engr)
```

## Ambiguous trailing zeros

Subset the data to look at just the numerical variables. 

```{r}
x <- density %>%
  select(T_K, p_Pa, R, density)
```

Print the data with incrementally decreasing significant digits. 

```{r}
kable(format_engr(x, sigdig = 4), caption = "4 digits")
```

Three digits creates no ambiguity. 

```{r}
kable(format_engr(x, sigdig = 3), caption = "3 digits")
```

With 2 digits, we have three columns with ambiguous trailing zeros. 

```{r}
kable(format_engr(x, sigdig = 2), caption = "2 digits")
```

By setting the `ambig_0_adj` argument to TRUE, scientific notation is used to remove the ambiguity. 

```{r}
kable(format_engr(x, sigdig = 2, ambig_0_adj = TRUE), caption = "Removing ambiguity")
```

The ambiguous trailing zero adjustment is applied only to those variables for which the condition exists. For example, if the pressure were known to only 2 digits, it is the only variable with ambiguous trailing zeros. 

```{r}
kable(format_engr(x, sigdig = c(4, 2, 0, 3), ambig_0_adj = FALSE))
```

With `ambig_0_adj = TRUE`, only the pressure variable has a reformatted power of ten. 


```{r}
kable(format_engr(x, sigdig = c(4, 2, 0, 3), ambig_0_adj = TRUE))
```

## align_pander() 

This function uses `pander()` to print a table and `panderOptions('table.alignment.default')` to align columns.  Usage is: `align_pander(x, align_idx = NULL, caption = NULL)` 

- `x` is the data frame 
- `align_idx` is a string comprised of any combination of  "r", "l", and "c"  The default alignments are numeric right and everything else left 
- `caption` is an optional string used as the `pander()` caption argument 

```{r}
align_pander(density_engr)
```

Finally, the heading can be edited for presentation. 

```{r}
names(density_engr) <- c(
  "Date",
  "Trial",
  "Temp (K)",
  "Press (Pa)",
  "R (J kg^-1^K^-1^)",
  "$\\rho$ (kg/m^3^)"
)
```

- backslashes have to be escaped, hence `\\rho` 
- superscripts like m^3^ in R Markdown are produced with `m^3^`


```{r}
align_pander(density_engr, "lccccc", caption = "Air density measurements")
```

## Conclusion 

These two functions provide the means for consistently rendering numbers with the desired number of significant digits, including trailing zeros, and align them in output tables without affecting character data in the same data frame. 




