---
title: "Numbers in engineering format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Numbers in engineering format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates the use of two functions from the *docxtools* package: 

- `format_engr()` for formatting numbers in engineering notation  
- `align_pander()` for aligning table columns using a "simple" pander table style 

## format_engr()

The primary goal of `format_engr()` is to present numeric variables in a data frame in *engineering format*, that is, scientific notation with exponents that are multiples of 3. Compare: 

| Syntax |Expression|
|--------|----------|
|conventional computer syntax |$1.011e+5$         | 
|mathematical syntax          |$1.011\times10^{5}$|
|engineering format           |$101.1\times10^{3}$|

This example uses a small temperature-pressure data set to compute air density and display the results in a table. Density is computed using the ideal gas law, $\rho = p / (RT)$. 


Start by loading packages, 


```{r}
library(knitr)
opts_knit$set(root.dir = "../")
suppressPackageStartupMessages(library(dplyr))
library(docxtools)
```

Create some data for the example,  

```{r}
# temperature in K
T_K  <- c(294.05, 294.15, 294.65, 293.35, 293.85)
# convert pressure from hPa to Pa
p_Pa <- c(1011, 1010, 1011, 1010, 1011) * 100
# gas constant in J / (kg K)
R    <- 287
# density in kg / m^3
density_data <- data.frame(T_K, p_Pa, R)
```

Compute air density for each observation,  

```{r}
density_data <-	mutate(density_data, density = p_Pa / (R * T_K))
knitr::kable(density_data)
```

The first argument of `format_engr()` is the data frame, the second is the array of significant digits in column order.  

```{r}
engr_density_data <- format_engr(density_data, sigdig = c(5, 4, 0, 5))
```

Data frame values are returned as character strings with math formatting. By setting `sigdig = 0` for **R**, the third column will be displayed in its  original form. 

```{r, collapse = TRUE}
str(engr_density_data)
```

The math formatting is applied when the data frame is printed in the output document.   

```{r}
knitr::kable(engr_density_data)
```

Comments: 

- scientific notation is omitted when the exponent is 0, 1, or 2, hence temperature **T_K** is reported with 5 digits but without scientific notation 
- pressure data **p_Pa** is correctly shown in engineering format 
- trailing zeros are significant 


## align_pander() 

This function uses `pander()` to print a table and `panderOptions('table.alignment.default')` to align columns.  Usage is: `align_pander(x, align_idx = NULL, caption = NULL)` 

- `x` is the data frame 
- `align_idx` is a string comprised of any combination of  "r", "l", and "c"  The default alignments are numeric right and everything else left 
- `caption` is an optional string used as the `pander()` caption argument 

```{r}
align_pander(engr_density_data, align_idx = "cccc")
```

Finally, the heading can be edited for presentation. The heading row, like the numeric variables, are formatted in R Markdown math format.

```{r}
names(engr_density_data) <- c("$T\\text{ (K)}$"
	, "$p\\text{ (Pa)}$"
	, "$R\\text{ (J kg}^{-1}\\text{ K}^{-1}\\text{)}$"
	, "$\\rho\\text{ (kg/m}^{3}\\text{)}$"
	)
```

- backslashes have to be escaped, hence `\\text{}` and `\\rho` 
- `\text{}` produces "roman" (non-italic) type inside the math environment 

```{r}
align_pander(engr_density_data, "cccc", caption = "Air density measurements")
```






## non-numeric variables  

Create some alphanumeric test data, 

```{r}
# create test input arguments
set.seed(20161221)
n  <- 5
a  <- sample(letters, n)
b  <- sample(letters, n)
x  <- runif(n, min =  -5, max =  50) * 1e+5
y  <- runif(n, min = -25, max =  40) / 1e+3
z  <- runif(n, min =  -5, max = 100)
alpha_num <- data.frame(z, b, y, a, x, stringsAsFactors = FALSE)
```

Format the entire data frame with the default 4 significant digits. 

```{r}
engr_alpha_num <- format_engr(alpha_num)
align_pander(engr_alpha_num, "rcrcr")
```

- character variables `b` and `a` are unaffected by `engr_format()`  
- coefficients have floating decimal points 

Variables can be re-ordered in the usual way, e.g., 

```{r}
alpha_num <- select(alpha_num, a, b, x, y, z)
engr_alpha_num <- format_engr(alpha_num)
align_pander(engr_alpha_num, "ccrrr")
```

## significant zeros 

Leading zeros are not significant. 

Trailing zeros generally should be significant. For example, isolate a number from the `y` column:  

```{r collapse = TRUE}
y2 <- alpha_num$y[2]
y2
```

Formatting `y2` with different significant digits using `format_engr(y2, sigdig)` yields the table below. With 3 digits, `y2` has 3 unambiguous significant digits. However, reducing the number of digits to 2 would produce a coefficient of $350$ with an ambiguous zero before the decimal point. In such cases, `format_engr()` changes the exponent to produce `r format_engr(y2, 2)` with two unambiguous significant digits.   

|`sigdig`|$y_2$|
|--------|----------|
|   4    |`r format_engr(y2, 4)`|
|   3    |`r format_engr(y2, 3)`|
|   2    |`r format_engr(y2, 2)`|
|   1    |`r format_engr(y2, 1)`|

*Exceptions to the significant trailing zero rule.* Consider $z_2$ from the data frame, 

```{r collapse = TRUE}
z2 <- alpha_num$z[2]
z2
```

For numbers like `z2` that in scientific notation would have exponents = 0, 1, or 2, `format_engr()` foregoes powers of ten notation.  


|`sigdig`|$z_2$|
|--------|----------|
|   4    |`r format_engr(z2, 4)`|
|   3    |`r format_engr(z2, 3)`|
|   2    |`r format_engr(z2, 2)`|
|   1    |`r format_engr(z2, 1)`|

As we reduce the number of significant digits, we can eventually obtain a trailing zero whose significance is ambiguous, as in $z_2 =$ `r format_engr(z2, 1)`. In such cases, `format_engr()` leaves the ambiguous significant zero instead of imposing scientific notation that the audience might find distracting.

## scalars and vectors are returned as data frames 

The preferred input to `format_engr()` is a data frame. If the input is a numeric vector, it will be formatted and returned as a data frame with the variable name `value`. 

For example,  using `y2` again, 

```{r, collapse = TRUE}
str(y2)
```

Now formatting the number returns a data frame with one row and column. 

```{r, collapse = TRUE}
engr_y2 <- format_engr(y2, 4)
str(engr_y2)
```

which can be used in an inline code chunk, e.g., 

    $y_2 =$ `r engr_y2$value`

to produce $y_2 =$ `r engr_y2$value`. 


A numeric vector is also returned as a data frame. 

```{r, collapse = TRUE}
# the x array
cat(x, sep = "\n")
class(x)

# formatted 
engr_x <- format_engr(x, 3)
class(engr_x)
engr_x
```

The input to `format_engr()` must have at least one numeric variable, or an error is thrown, e.g., running `format_engr(a)` produces the error 

`Error: m_numeric_cols > 0 is not TRUE Execution halted`. 

## conclusion 

These two functions provide the means for consistently rendering numbers with the desired number of significant digits, including trailing zeros, and align them in output tables without affecting character data in the same data frame. 




